100 years of Pies vs. Bars

### 2025-04-07 Draft.qmd
updated

# 4. Criticism of previous experimental design

>>>>

# 4. Criticism of previous Experimental design

## 4.1 Experiment Description

This poster synthesizes findings from a series of historical and modern experiments comparing pie and bar charts for interpreting proportional data. Early studies, such as Eells (1926) and Croxton & von Huhn (1927), focused on accuracy and speed in estimating single or dual category proportions (e.g., A vs. B or A / (A + B)). These experiments typically involved participants identifying which portion of a chart was larger or estimating ratios between segments.

Modern experiments, such as Spence & Lewandowsky (1991), expanded the task complexity by asking participants to compare single components (A vs. B), a component against a pair (A vs. B + C), and two pairs (A + B vs. C + D). These tasks were tested across multiple chart formats (pie, bar, and table), with variations in exposure time, number of components, and use of color or ordering. Performance was evaluated in terms of speed and accuracy, allowing researchers to assess how perceptual and cognitive demands shift across visualization types and task difficulty.

Peterson and Schramm (1954) evaluated the accuracy of proportion estimation across eight common chart types used to represent parts of a whole: circle, disc, single bar, multiple bar, multiple cylinder, multiple square column, multiple area column, and partial cosmograph. Each chart displayed five fixed proportions (43%, 23%, 18%, 10%, 6%) and was presented alongside a similar “decoy” graph to mask repetition. A total of 112 U.S. Air Force personnel viewed the charts in random order and estimated the percentage values for each component. After discarding incomplete responses, 86 participants contributed 3,440 data points, forming the basis for comparison across visual formats.

Cleveland & McGill (1984) conducted a series of experiments to test a theory of graphical perception based on “elementary perceptual tasks” such as position, length, angle, area, and shading. They hypothesized a hierarchy of these tasks based on how accurately people extract quantitative information from them. In their experiments, participants viewed a range of common graph types—including bar charts, pie charts, and divided bar charts—and were asked to make visual judgments of values or differences. The results supported their theory: judgments involving position along a common scale were the most accurate, followed by length, with angle (used in pie charts) yielding the least accurate responses. These findings laid the foundation for many modern principles in data visualization design.

Cleveland & McGill (1986) conducted a foundational experiment to evaluate the accuracy of seven basic graphical perception tasks: position (aligned and non-aligned), length, angle, slope, and area (circle and blob). Participants were shown visual stimuli in which one object served as a standard, and three comparison objects were judged as percentages relative to it. Each participant made 210 judgments across 70 displays. The study included 127 participants from diverse educational backgrounds. Results confirmed a perceptual hierarchy: position judgments were the most accurate, while angle and area were less reliable—insights that continue to influence data visualization design today.

Simkin and Hastie (1987) investigated how different graphical encodings—position, length, and angle—affect performance on visual judgment tasks. 200 Undergraduate participants were shown pie charts, bar charts, and divided bar charts and asked to make either comparison judgments (A/B) or proportion-of-the-whole judgments. Each graph type relied on a different visual code: bar charts used position, divided bars used length, and pie charts used angle. Participants completed 90 trials under timed conditions. Results showed that position and length led to higher accuracy and faster response times for comparison tasks, while pie charts supported better performance in proportion-of-the-whole judgments, demonstrating a clear interaction between graph type and task type.

Very recently, more works have been discovered on this debate that catch researchers’ eyes on the topic. Kosara & Skau (2016) investigated how common distortions in pie charts affect judgment accuracy. Participants were shown a series of pie chart variations and asked to estimate the percentage represented by a highlighted slice. The variations included a standard pie chart, an "exploded" pie (with a slice offset), a "larger slice" chart (with an extended radius), a vertically compressed elliptical pie, and a square-shaped chart. All charts had two segments and were rotated randomly to eliminate positional bias. Participants were recruited via Mechanical Turk and asked to provide whole-number estimates for values ranging from 3% to 97%. Results showed that all distortions—except the exploded pie—led to significantly higher judgment errors compared to the baseline pie chart.

In 2025, Hill did the study that involved three online experiments designed to evaluate how effectively bar, pie, and donut charts support part-whole estimation and ranking tasks. In Experiment 1, participants viewed randomized sets of charts (10 bar, 5 pie, 5 donut) created from multinomially generated data and answered ranking and proportion estimation questions. Experiment 2 focused on charts where elements A and B were similar in size, allowing for precise comparison of visual discrimination accuracy across chart types. Experiment 3 assessed relative magnitude estimation on pie and donut charts using a direct comparison task (“If A is 100, what is B?”). A total of 42 participants from the University of Surrey took part in the study via Qualtrics, and all data were anonymized

## 4.2 Criticism of previous Experimental design
